---
title: "Building and Applying a HOMED Reference"
output:
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Building and Applying a HOMED Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

HOMED (Hierarchically Optimized Methylation Deconvolution) is a supervised, reference-based framework for estimating cell-type composition from bulk DNA methylation (DNAm) data. The method leverages purified cell-type DNAm profiles together with training samples containing known or pseudo-ground-truth cell-type proportions to construct optimized methylation reference matrices. HOMED first identifies informative CpG sites through differential methylation analysis and then refines the resulting reference libraries using an iterative IDOL-based optimization procedure that minimizes deconvolution error on training samples. To account for biological relationships among cell populations, HOMED supports a two-level hierarchical structure in which broad lineage-level cell types can be further subdivided into finer subtypes. During deconvolution, cell-type proportions are estimated hierarchically using constrained projection, and subtype abundances are rescaled to ensure consistency with their parent lineages. The resulting framework provides accurate and biologically coherent estimates of cell-type composition from bulk DNAm data.

This vignette demonstrates the HOMED workflow using a PBMC example dataset included with the package.

The typical workflow consists of:

1. Data preparation and scRNA-seq-based deconvolution to obtain ground truth
2. Constructing a hierarchical HOMED reference
3. Estimating cell-type proportions from bulk methylation data
4. Evaluating deconvolution performance

# Data preparation and scRNA-seq-based deconvolution
For this vignette, we assembled a PBMC benchmark dataset consisting of purified methylation references, matched bulk gene expression and methylation profiles, and a scRNA-seq reference atlas. Bulk expression and matched DNA methylation samples were assembled from three public PBMC studies (GSE105124, GSE49065, and GSE40736). 

Expression data from different microarray platforms were mapped to gene symbols using platform-specific annotations and collapsed to gene-level measurements, while methylation data were converted to beta values when necessary. Samples with both expression and methylation profiles available were matched by sample identifier and combined into a unified PBMC bulk dataset before being randomly split into training (70%) and testing (30%) cohorts. 

Purified PBMC methylation reference profiles were obtained from GSE167998. Cell types were harmonized into a hierarchical structure consisting of four level-1 populations (T cells, B cells, NK cells, and monocytes) and five level-2 populations (CD4+ T cells, CD8+ T cells, B cells, NK cells, and monocytes). Specifically, CD4nv, CD4mem, CD4, and Treg samples were grouped as CD4+ T cells; CD8nv and CD8mem samples were grouped as CD8+ T cells; Bnv and Bmem samples were grouped as B cells; monocytes and NK cells were kept as is. 

A PBMC scRNA-seq atlas from 10x Genomics was used as the expression reference for DWLS deconvolution. Single-cell annotations were harmonized into five target cell types matching the methylation reference. Specifically, CD4, CD4mem, CD4nv, and Treg cells were grouped as CD4+ T cells; CD8mem and CD8nv cells were grouped as CD8+ T cells; B cells, NK cells, and monocytes were kept as is. CD34+ progenitor cells were excluded from downstream analysis. DWLS deconvolution was then applied to the matched bulk gene expression samples using this reference to generate pseudo-ground-truth cell-type proportions. These pseudo-ground-truth proportions of the training set were subsequently used to optimize the HOMED methylation reference matrix, and the proportions of the testing set were used for validation for prediction accuracy.

For convenience, the fully processed datasets used in this tutorial are hosted on Hugging Face. Users can download them here:

```{r}
download.file(
  "https://huggingface.co/datasets/Durianhead/HOMED_data/resolve/main/PBMC_data.tar.gz",
  destfile = "PBMC_data.tar.gz",
  mode = "wb"
)

untar("PBMC_data.tar.gz")
```

# Construct a Hierarchical Reference

The `HOMED_Reference()` function constructs an optimized hierarchical DNA methylation reference using purified cell-type methylation profiles, cell-type annotations, training bulk methylation samples, and corresponding cell-type proportions. These cell-type proportions serve as ground truth during reference optimization. In this example, pseudo-ground-truth proportions were generated using DWLS RNA-based deconvolution by combining matched bulk RNA-seq samples with a PBMC single-cell RNA-seq reference atlas. 

```{r load_data}
library(HOMED)

pure_methyl <- readRDS("PBMC_pure_methyl.rds")
pure_celltype <- readRDS("PBMC_pure_methyl_pdata.rds")
bulk_methyl_train <- readRDS("PBMC_bulk_methyl_train.rds")
bulk_cellprop_train <- readRDS("PBMC_bulk_cellprop_train.rds")

dim(pure_methyl)
dim(pure_celltype)
dim(bulk_methyl_train)
dim(bulk_cellprop_train)
```

HOMED also requires a user-defined cell-type hierarchy. For PBMCs, we use a two-level hierarchy in which CD4+ and CD8+ T cells are grouped under a common parent T-cell lineage, while monocytes, B cells, and NK cells are treated as independent cell types. Note that calling `HOMED_Reference()` might take around 20 minutes.

```{r}
two_level_celltypes <- list(
    mono = NULL,
    t = c("cd4t", "cd8t"),
    nk = NULL,
    b = NULL
)

ref <- HOMED_Reference(
    FACS_beta = pure_methyl, 
    FACS_pd = pure_celltype, 
    training_beta = bulk_methyl_train, 
    training_prop = bulk_cellprop_train, 
    celltypes = two_level_celltypes
)
```

# Estimating cell-type proportions from bulk methylation data

Once the reference matrix has been constructed, it can be supplied to `HOMED_Estimate()` together with a new set of bulk DNA methylation samples. `HOMED_Estimate()` uses the optimized hierarchical reference to infer the cell-type composition of each bulk sample, returning estimated proportions for all cell types defined in the hierarchy. In this example, the bulk methylation samples used for deconvolution are a held-out test set derived from the same PBMC dataset as the training samples. These samples were not used during reference construction and therefore provide an independent evaluation of HOMED's deconvolution performance.

```{r estimate}
bulk_methyl_test <- readRDS("PBMC_bulk_methyl_test.rds")

est_bulk_test <- HOMED_Estimate(
    celltypes = two_level_celltypes,
    deconv_beta = bulk_methyl_test,
    reference_obj = ref
)$deconvolution_proportion
```

# Evaluating deconvolution performance

To evaluate the performance of HOMED, we compare the estimated proportions of the held-out test set with the pseudo-ground truth generated by DWLS.

```{r}
meta_test <- readRDS("PBMC_bulk_cellprop_test.rds")

cts <- intersect(colnames(meta_test), colnames(est_bulk_test))

data.frame(
    CellType = cts,
    PCC = sapply(
        cts,
        function(ct) cor(meta_test[,ct], est_bulk_test[,ct])
    ),
    RMSE = sapply(
        cts,
        function(ct)
            sqrt(mean((meta_test[,ct]-est_bulk_test[,ct])^2))
    )
)
```
To visualize deconvolution performance, we generate scatter plots comparing the estimated cell-type proportions against the pseudo-ground-truth proportions for each PBMC cell type. Each point represents one bulk sample, and the red diagonal line indicates perfect agreement between the estimated and reference proportions. Higher Pearson correlation coefficients (PCC), lower root mean squared errors (RMSE), and points clustering closely around the diagonal indicate more accurate cell-type proportion estimates.

```{r fig.width=12, fig.height=3}
meta_test <- readRDS("PBMC_bulk_cellprop_test.rds")

cts <- c("mono","b","cd8t","nk","cd4t")

par(mfrow = c(1,length(cts)), mar = c(4,4,2,1))

lims <- range(
    c(meta_test[,cts], est_bulk_test[,cts]),
    na.rm = TRUE
)

for(ct in cts){

    truth <- meta_test[,ct]
    est <- est_bulk_test[,ct]

    plot(
        truth,
        est,
        main = ct,
        xlab = "True",
        ylab = "Estimated",
        pch = 16,
        xlim = lims,
        ylim = lims,
        asp = 1
    )

    abline(0,1,col="red",lwd=2)

    legend(
        "topleft",
        bty = "n",
        cex = 0.8,
        legend = c(
            paste0("PCC=", round(cor(truth,est),2)),
            paste0("RMSE=", round(sqrt(mean((truth-est)^2)),2))
        )
    )
}
```
